Spike spectra for recurrences

### 2022-10-31 Manuscript/inter_spike_manuscript_mdpi.tex
Reviewer2 more refs on other decomposition techniques

For example, consider the signal we would like to analyze 
(e.g., the $\tau$-RR of a system) to be a Dirac comb (DC) with inter-spike period $T_\text{is}$: 
\begin{equation}
\text{DC}_{\text{is}}(t) = \sum_{k=-\infty}^{\infty} \delta(t-kT_\text{is}),
\label{eq_dirac_comb}
\end{equation}
i.e., a series of Dirac delta functions for a period $T_\text{is}$. There is only one single period, $T_\text{is}$, in this signal (Fig.~\ref{fig_tau_rr_dirac_comb}A, D), so 
in principle we would strive for a single peak in the frequency domain of this signal at a frequency $f=1/T_\text{is}$. The Fourier spectrum does not meet this expectation 
and instead of a single frequency, there are many frequencies excited (Fig.~\ref{fig_tau_rr_dirac_comb}B, E). This is because the Fourier components add constructively 
for every frequency $1/T_\text{is}$ and therefore $\text{DC}_{\text{is}}(t)$ coincides with its own Fourier transform up to a factor $1/T_\text{is}$.\footnote{This can also 
be observed for neuron spike trains, e.g. \cite{Orcioni2020,Biagetti2017}.}

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea. Note that this approach can be used to decompose arbitrary signals, and is not specifc to $\tau$-RR. However, the more spiky the signal is, the more useful our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

The signal, which we would like to transform is decomposed into a set of appropriate basis functions.\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} Instead of using trigonometric 
functions, as it is the idea in the Fourier decomposition, we use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

>>>>

For example, consider the signal we would like to analyze 
(e.g., the $\tau$-RR of a system) to be a Dirac comb (DC) with inter-spike period $T_\text{is}$: 
\begin{equation}
\text{DC}_{\text{is}}(t) = \sum_{k=-\infty}^{\infty} \delta(t-kT_\text{is}),
\label{eq_dirac_comb}
\end{equation}
i.e., a series of Dirac delta functions for a period $T_\text{is}$. There is only one single period, $T_\text{is}$, in this signal (Fig.~\ref{fig_tau_rr_dirac_comb}A, D), so 
in principle we would strive for a single peak in the frequency domain of this signal at a frequency $f=1/T_\text{is}$. The Fourier spectrum does not meet this expectation 
and instead of a single frequency, there are many frequencies excited (Fig.~\ref{fig_tau_rr_dirac_comb}B, E). This is because the Fourier components add constructively 
for every frequency $1/T_\text{is}$ and therefore $\text{DC}_{\text{is}}(t)$ coincides with its own Fourier transform up to a factor $1/T_\text{is}$ \cite{Norden1998}.\footnote{This can also 
be observed for neuron spike trains, e.g. \cite{Orcioni2020,Biagetti2017}.}

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea. Note that this approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. However, the more spiky the signal is, the more useful our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

The signal, which we would like to transform\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} is decomposed into a set of appropriate basis functions. 
The general idea is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. 
A number of decomposition techniques are out there based on different sets of basis functions, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (Principal Component Analysis \cite{Hotelling1933} and related techniques) 
or Intrinsic Mode Functions (Empirical Mode Decomposition and Hilbert spectrum \cite{Norden1998}).

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}
